Supplement 2: Stationwise Evaluation

### 2018-09-23 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
enhancements in stationwise supplement
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2295 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 80df211ca44fd0cbd1a862b185b25559c2c1615b

\Abstract{In the application discussed in the paper ''Distributional 
Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
the presented method is evaluated on 95 observation stations. More extensive
evaluations including a cross-validation are preformed for one station, Axams.
In this supplement this exact same analysis as for station Axams is
made for 14 further observation stations.}

@@

\begin{document}
\SweaveOpts{concordance=TRUE}

\section{Introduction}
This application study is supplementary material to the application in Section~3 of 
the paper where the presented distributional forest model is employed to obtain 
probabilistic precipitation forecasts in a mountainous region.
Next to an evaluation over all 95 observation stations further investigations
including a cross-validation have been performed for one selected station, Axams.
Here, this more extensive evaluation is carried out, applying the exact same settings 
as in the paper, for Axams and 14 further observation stations. Additionally, computation
times are measured.

\section{Models and evaluation}
As in the paper a distributional forest applying a left-censored at zero 
Gaussian distribution is evaluated together with three other zero-censored 
Gaussian models. Further details on all four models can be found in Section~3.2. of
the paper. In particular, Table~2 gives an overview over the models and their 
specifications. As a brief reminder the approaches are summarized below.
%
\begin{itemize}

@@

\item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
Standard ensemble model output statistics models use the ensemble 
mean of total precipitation as regressor in the location submodel 
and the corresponding ensemble standard deviation in the scale submodel.
The parameters are estimated by maximum likelihood.

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
the most relevant predictors based on meteorological expert knowledge
following \cite{Stauffer+Umlauf+Messner:2017}. The model is estimated by maximum
penalized likelihood using a backfitting algorithm (\cite{Stasinopoulos+Rigby:2007}).

\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
automatically from all available variables, using non-cyclic boosting
for parameter estimation (\cite{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}).
The optimal stopping point for the iteration is evaluated based on 
a (computationally intensive) out-of-bag bootstrap estimate of the log-likelihood.

@@

%\subsection{Subset selection}
The models are always learned on data from 24 years and tested on the 
remaining 4 years. Splitting the data into these two subsets is done
in two different ways leading to the two following evaluation settings.
%
\begin{itemize}

\item Single split: The data is time-ordered. The learning data consists of the first 24 years (1985--2008) and the testing data of the following 4 years (2009--2012).

\item Cross-validation: In order to not just split the data at one splitpoint the 28 years are randomly split in 7 blocks of 4 years which are not necessarily successive. Each of the blocks is chosen once as testing data while the remaining 6 blocks are the learning data. This procedure is repeated 10 times resulting in a 10 times 7-fold cross-validation framework.

@@

The performances of the models are compared based on numerical and graphical evaluations.

\textit{Numerical:}
For the single split setting one value is calculated for each observation in the test data and the mean of these values is considered. For the cross-validation this mean value over all observations in the test data is retrieved from all 70 evaluations and then again the mean of them is considered. This holds for the following measurements.
%
\begin{itemize}

\item CRPS: The continuous ranked probability score (CRPS) is calculated to assess the predictive performance in terms of full probabilistic forecasts.

\item Time: For all models fitting times and predictions times are measured. In particular, \textit{fit time} refers to the time needed to fit the model to the whole learn data. Depending on the setting \textit{predict time} refers to the time needed to make one prediction in the single split setting and to make predictions for the whole test data in the cross-validation.

\textit{Graphical:} The following methods of visualization are applied to provide further means of comparison of all models but also to investigate on the distributional forest in particular.
%
\begin{itemize}
%
\item Boxplot of the CRPS skill score: With the EMOS as reference model the CRPS skill scores are calculated for the cross-validation setting and illustrated in a boxplot.
%
\item Barplot of variable importance: For the forest model it is investigated which variables have the strongest influence on the model and its goodness of fit. For each variable this is measured based on mean decrease in CRPS after permutation. The 10 variables showing the highest values in the single split setting are listed in a barplot.
%
\item Residual QQ plots: For each of the models an out-of-sample residual QQ plot for the single split setting provides a visualization of goodness of fit. Randomized residual quantiles are compared to standard normal quantiles.
%
\item PIT histograms: As an additional visualization out-of-sample probability integral transform histograms illustrate how well-calibrated the distributional fits are for the single split setting.
%
\end{itemize}

\section{Stations}
Out of the 95 observation stations 15 are considered for this 
benchmark study. They have been selected based on the region
they are in, their altitude and the performance of the 
different models in the application in the paper such that
all of these features are represented equally.

Selected stations: \hyperlink{sumAxams.1}{Axams},
\hyperlink{sumLech.1}{Lech},
\hyperlink{sumZuers.1}{Zuers},
\hyperlink{sumSeeimPaznaun.1}{See im Paznaun},
\hyperlink{sumJungholz.1}{Jungholz},
\hyperlink{sumLadisNeuegg.1}{Ladis-Neuegg},
\hyperlink{sumOetz.1}{Oetz},
\hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut},
\hyperlink{sumGinzling.1}{Ginzling},
\hyperlink{sumRotholz.1}{Rotholz},
\hyperlink{sumWalchsee.1}{Walchsee},
\hyperlink{sumKoessen.1}{Koessen},
\hyperlink{sumInnervillgraten.1}{Innervillgraten},
\hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol},
\hyperlink{sumStJohannimWalde.1}{St.Johann im Walde}.

@@

\subsection{Forecasting skills} 
Comparing CRPS values of the single split setting it can be stated that the forest model works well over all stations. Often it is even the best performing method (e.g. at stations \hyperlink{sumLech.1}{Lech}, \hyperlink{sumSeeimPaznaun.1}{See im Paznaun} and \hyperlink{sumGinzling.1}{Ginzling}), but there are also stations where one of the other models leads to better results. For example at \hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut} the boosted GAMLSS performs best, at \hyperlink{sumRotholz.1}{Rotholz} the EMOS model and at \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol} the prespecified GAMLSS is the best model, however, they do not clearly outperform the forest model. Moreover, these advantages often even out or turn into disadvantages when looking at the CRPS values from the cross-validation. For example at station \hyperlink{sumInnervillgraten.1}{Innervillgraten}, the prespecified GAMLSS is clearly the best in the single split setting, however, in the cross-validation the forest model is slightly ahead. An exception is given by station \hyperlink{sumSt.JohannimWalde.1}{St.Johann im Walde} as it is the only station where the prespecified GAMLSS clearly outperforms all other models in both settings. But overall, the results from the cross-validation also support the statement that the forest model performs well on all stations being either the best or among the best performing methods at the majority of stations.
Moreover, it can be concluded that Axams is an observation station without any unusual features that would lead to outstanding performances of any of the methods. Therefore, it is a reasonable choice as a representative for all 95 station. Examples for stations where the forest model performs better or worse can be found easily, however, in general the methods perform similarly as for station Axams.

%Comparing CRPS skill score values from the cross-validation the distributional forest
%and the two GAMLSS versions show clear improvements compared to the reference model EMOS.
%At many stations the forest and either the prespecified GAMLSS (e.g. at station Innervillgraten) or the boosted GAMLSS (e.g. at station Lech or Zuers) are ahead, almost on the same level, while the second GAMLSS is slightly behind. At stations Axams, See im Paznaun and Ginzling the forest clearly leads to the highest improvement in terms of CRPS. Station St.Johann im Walde stands out as it is the only of the 15 stations where the prespecified GAMLSS is clearly ahead of the other models. But overall it can be stated that all three models perform well with only small variations over stations. Even if one of the evaluated models shows clear advantages in the single split setting (such as the EMOS at station Rotholz or the prespecified GAMLSS at station Innervillgraten) this is evened out within the cross-validation.

\subsection{Computation time}
The forest model and the two GAMLSS models achieve similar results in both time measurements, fit and predict time, while the EMOS model is clearly the fastest. However, for the boosted GAMLSS the cross-validation applied to find the optimal stopping value \code{mstop} is extremely time consuming. This additional time can not be neglected since the stopping value is crucial for the complexity of the model and the selection of covariates.

As to prediction times it has to be mentioned that the forest can keep up with the other models if only one prediction is made. However, to make predictions for a whole set of new observations the forest clearly needs more time. This is due to the fact that the forest has to computes each single prediction separately as the maximum-likelihood estimation is based on observation-specific weight matrix.

\subsection{Variable importance}
Concerning variable importance, two covariates based on total precipitation (\code{tp\_max} and \code{tp\_mean}) are at the top of the list for all stations, often clearly ahead of the other covariates. The other members of the top ten variables vary slightly over stations, however, further covariates concerning total precipitation, total column-integrated condensate and temperature differences are often among them.

\subsection{Calibration}
The PIT histograms and residual QQ plots show that overall all four models are reasonably well calibrated at the majority of stations while at some stations they all fit better (e.g. at station \hyperlink{sumOetz.1}{Oetz}) or worse (e.g. at station \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol}) than on average. One feature of the boosting method that can be detected in the residual QQ plots for many stations is that heavy upper tails are not captured very well. This is also supported by the PIT histogram.

\textit{Note:} As mentioned by \cite{Hofner+Mayr+Schmid:2016} the methods provided by GAMLSS ``can be unstable, especially when it comes to selecting possibly different sets of variables for multiple distribution parameters''. These problems have also occurred in a few situations for the 
prespecified GAMLSS model within the cross-validation. In case of this method failing to 
build a model it is represented by \code{NA}s in the evaluation.

\includepdf[pages=-, link, linkname=sumAxams]{Axams.pdf}
\bookmark[dest=sumAxams.1]{Axams}
\includepdf[pages=-, link, linkname=sumLech]{Lech.pdf}
\bookmark[dest=sumLech.1]{Lech}
\includepdf[pages=-, link, linkname=sumZuers]{Zuers.pdf}
\bookmark[dest=sumZuers.1]{Zuers}
\includepdf[pages=-, link, linkname=sumSeeimPaznaun]{SeeimPaznaun.pdf}
\bookmark[dest=sumSeeimPaznaun.1]{See im Paznaun}
\includepdf[pages=-, link, linkname=sumJungholz]{Jungholz.pdf}
\bookmark[dest=sumJungholz.1]{Jungholz}
\includepdf[pages=-, link, linkname=sumLadisNeuegg]{LadisNeuegg.pdf}
\bookmark[dest=sumLadisNeuegg.1]{Ladis-Neuegg}
\includepdf[pages=-, link, linkname=sumOetz]{Oetz.pdf}
\bookmark[dest=sumOetz.1]{Oetz}
\includepdf[pages=-, link, linkname=sumOchsengartenObergut]{OchsengartenObergut.pdf}
\bookmark[dest=sumOchsengartenObergut.1]{Ochsengarten-Obergut}
\includepdf[pages=-, link, linkname=sumGinzling]{Ginzling.pdf}
\bookmark[dest=sumGinzling.1]{Ginzling}
\includepdf[pages=-, link, linkname=sumRotholz]{Rotholz.pdf}
\bookmark[dest=sumRotholz.1]{Rotholz}
\includepdf[pages=-, link, linkname=sumWalchsee]{Walchsee.pdf}
\bookmark[dest=sumWalchsee.1]{Walchsee}
\includepdf[pages=-, link, linkname=sumKoessen]{Koessen.pdf}
\bookmark[dest=sumKoessen.1]{Koessen}
\includepdf[pages=-, link, linkname=sumInnervillgraten]{Innervillgraten.pdf}
\bookmark[dest=sumInnervillgraten.1]{Innervillgraten}
\includepdf[pages=-, link, linkname=sumMatreiinOsttirol]{MatreiinOsttirol.pdf}
\bookmark[dest=sumMatreiinOsttirol.1]{Matrei in Osttirol}
\includepdf[pages=-, link, linkname=sumStJohannimWalde]{StJohannimWalde.pdf}
\bookmark[dest=sumStJohannimWalde.1]{St.Johann im Walde}

>>>>

\Abstract{
  In the case study presented in the main manuscript ``Distributional Regression
  Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
  the novel distributional regression forests are evaluated and compared with
  other distributional regression approaches on 95 meteorological stations.
  Moreover, further details -- including a cross-validation analysis -- are
  preformed for only one station, Axams. To show that Axams is a fairly typical
  station and similar insights can be obtained for other stations as well,
  this supplements presents the same analysis as in Section~3.3 of the main
  manuscript for 14~further meteorological stations.
}

@@

\begin{document}

\section{Setup}

\subsection{Models}

As introduced in Section~3.2 of the main manuscript, a distributional forest
based on a Gaussian distribution, left-censored at zero, is evaluated together
with three other zero-censored Gaussian models. Table~2 of the manuscript provides
an overview of all models considered and their specifications. As a reminder
the approaches are briefly summarized below.
%
\begin{itemize}

@@

\item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
Standard ensemble model output statistics uses the ensemble 
mean of total precipitation as regressor in the location submodel 
and the corresponding ensemble standard deviation in the scale submodel.
The parameters are estimated by maximum likelihood.

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
the most relevant predictors based on meteorological expert knowledge
following \cite{Stauffer+Umlauf+Messner:2017}. The model is estimated by maximum
penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.

\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
automatically from all available variables, using non-cyclic boosting
for parameter estimation \citep{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}.
The optimal stopping point for the iterations is determined based on 
a (computationally intensive) out-of-bag bootstrap estimate of the log-likelihood.

@@

\subsection{Learning and test samples}

The models are always learned on data from 24 years and tested on the 
remaining 4 years. Splitting the data into these two subsets is done
in two different ways leading to the two following evaluation settings.
%
\begin{itemize}

\item \emph{Single split}: The data is time-ordered. The learning data consists of the
first 24 years (1985--2008) and the testing data of the following 4 years
(2009--2012).

\item \emph{Cross-validation}: In order to not just split the data at one splitpoint
the 28 years are randomly split into 7 blocks of 4 years which are not
necessarily successive. Each of the blocks is chosen once as testing data while
the remaining 6 blocks are the learning data. This procedure is repeated 10
times resulting in a 10 times 7-fold cross-validation framework.

\end{itemize}
%
The performances of the models are compared based on numerical and graphical evaluations.

\subsection{Numerical evaluation}

Several numerical summary statistics are computed in both
settings. For the single-split setting one value is calculated for each
observation in the test data and then their mean is reported. For the
cross-validation setting, the same is done for all 70~splits considered and
again their mean is reported. The following summary statistics are used.
%
\begin{itemize}

\item \emph{CRPS}: The continuous ranked probability score (CRPS) is calculated to
assess the predictive performance in terms of full probabilistic forecasts.

\item \emph{Computation time}: For all models fitting times and predictions times
are measured on the same 16-core Linux computer. In particular,
\textit{fit time} refers to the time needed to fit the model to the learning
data in a particular split. The \textit{predict time} refers to the time needed
to compute the out-of-sample prediction. In the single-split setting, the
predictions are computed observation by observation, i.e., emulating the
realistic setup where predictions are made once the new NWP model outputs become
available. In the cross-validation setting, the predictions for the full 4-year
out-of-sample period in one go. Although this is not realistic in this
particular application, it might be an interesting quantity for comparing the
competing regression approaches.

\end{itemize}
%
\textit{Note:} For a few splits in the cross-validation, the prespecified GAMLSS
(as computed by the \proglang{R} package \pkg{gamlss}) failed, leading to \code{NA}s
in the corresponding summary statistics. As mentioned by \cite{Hofner+Mayr+Schmid:2016}
such problems are not unusual because GAMLSS ``can be unstable, especially when it comes
to selecting possibly different sets of variables for multiple distribution parameters''.

\subsection{Graphical evaluation}

The following methods of visualization are applied to
provide further means of comparison of all models but also to investigate on the
distributional forest in particular.
%
\begin{itemize}

\item \emph{Boxplot of the CRPS skill score:} With EMOS as the reference model
the CRPS skill scores are calculated on each of the 70~splits considered in the
cross-validation setting and illustrated in a boxplot.

\item \emph{Barplot of variable importance:} For the distributional forest it is
investigated which variables have the strongest influence on the model and its
goodness of fit. For each variable this is measured based on mean decrease in
CRPS after permutation. The 10 variables showing the highest values in the
single-split setting are listed in a barplot.

\item \emph{Residual QQ plots:} For each of the models an out-of-sample residual
QQ plot for the single-split setting provides a graphical goodness-of-fit
assessment.

\item \emph{PIT histograms:} Similarly, the out-of-sample probability integral
transform (PIT) in the single-split setting is amployed as an alternative
goodness-of-fit visualization.

@@

\section{Stations}

Out of the 95~observation stations 15 are considered for this 
supplemental study. They have been selected to cover a wide range regarding
geographical location, altitude, and which of the competing models performed
best in the single-split setting reported in Section~3.4 of the main manuscript.

Selected stations: \hyperlink{sumAxams.1}{Axams},
\hyperlink{sumLech.1}{Lech},
\hyperlink{sumZuers.1}{Zuers},
\hyperlink{sumSeeimPaznaun.1}{See im Paznaun},
\hyperlink{sumJungholz.1}{Jungholz},
\hyperlink{sumLadisNeuegg.1}{Ladis-Neuegg},
\hyperlink{sumOetz.1}{Oetz},
\hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut},
\hyperlink{sumGinzling.1}{Ginzling},
\hyperlink{sumRotholz.1}{Rotholz},
\hyperlink{sumWalchsee.1}{Walchsee},
\hyperlink{sumKoessen.1}{Koessen},
\hyperlink{sumInnervillgraten.1}{Innervillgraten},
\hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol},
\hyperlink{sumStJohannimWalde.1}{St.~Johann im Walde}.

@@

\subsection{Forecasting skill}

Based on the out-of-sample CRPS results it can be concluded that the
distributional forest generally performs well on all stations and is typically
close to (e.g., \hyperlink{sumLech.1}{Lech}) or somewhat better than
the (next-)best method (e.g., \hyperlink{sumSeeimPaznaun.1}{See im Paznaun} and
\hyperlink{sumGinzling.1}{Ginzling}). Naturally, there is more variation
in the single-split results with another models performing better than the
distributional forest. However, in the cross-validation such differences
often even out, showing that distributional forests are on par. Examples for
this include \hyperlink{sumRotholz.1}{Rotholz},
\hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol}, and 
\hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut}, where the 
best models in the single split are EMOS, prespecified GAMLSS and boosted
GAMLSS, respectively. The only station, where another model performs
clearly better than distributional forests in both settings, is
\hyperlink{sumSt.JohannimWalde.1}{St.~Johann im Walde}.

Overall, the results support the conclusion that distributional forests
perform well for this prediction task -- they are often on par and sometimes
even better than the other models considered. Moreover, it is shown that the
results for Axams are fairly typical and not particularly unusual compared
to the other stations considered.

\subsection{Computation time}

Distributional forests are comparable with respect to the computation times
to the other two flexible GAMLSS models. Only the basic EMOS model is clearly
the fastest. Moreover, unlike the other models, the boosted GAMLSS requires
a time-consuming cross-validation to determine the optimal stopping value
\code{mstop}. This is crucial for selecting the appropriate model complexity
and relevant covariates and hence cannot be neglected in the analysis.

The observation-wise prediction times in the single-split setting are in the same
order of magnitude for the three flexible distributional models. However, when
many predictions are computed simultaneously in the cross-validation setting,
then distributional forests are slower than the GAMLSS models. The reason for this
is that the GAMLSS can set up the model matrices once and then carry out a matrix
product while the distributional forest necessitates fitting a separate set of
parameters for each observation-specific weight vector.

\subsection{Variable importance}

Two NWP model outputs based on total precipitation (\code{tp\_max} and
\code{tp\_mean}) are clearly the most important predictors for precipitation
at all stations considered. The remaining variables typically follow with
some margin and vary slightly in their exact order. However, they typically
encompass some NWP model outputs for total precipitation, total column-integrated
condensate, and temperature differences.

\subsection{Calibration}

The PIT histograms and residual QQ plots show that overall all four models are
reasonably well-calibrated at the majority of stations while at some stations
they all fit better (e.g., at station \hyperlink{sumOetz.1}{Oetz}) or worse (e.g.,
at station \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol}) than on
average. For several stations, however, the boosted GAMLSS does not capture
the heavy upper tail as well as the other methods which is best seen in the
residual QQ plots but also the PIT histograms.

\includepdf[pages=-, link, linkname=sumAxams]{Axams.pdf}
\bookmark[dest=sumAxams.1]{Axams}
\includepdf[pages=-, link, linkname=sumLech]{Lech.pdf}
\bookmark[dest=sumLech.1]{Lech}
\includepdf[pages=-, link, linkname=sumZuers]{Zuers.pdf}
\bookmark[dest=sumZuers.1]{Zuers}
\includepdf[pages=-, link, linkname=sumSeeimPaznaun]{SeeimPaznaun.pdf}
\bookmark[dest=sumSeeimPaznaun.1]{See im Paznaun}
\includepdf[pages=-, link, linkname=sumJungholz]{Jungholz.pdf}
\bookmark[dest=sumJungholz.1]{Jungholz}
\includepdf[pages=-, link, linkname=sumLadisNeuegg]{LadisNeuegg.pdf}
\bookmark[dest=sumLadisNeuegg.1]{Ladis-Neuegg}
\includepdf[pages=-, link, linkname=sumOetz]{Oetz.pdf}
\bookmark[dest=sumOetz.1]{Oetz}
\includepdf[pages=-, link, linkname=sumOchsengartenObergut]{OchsengartenObergut.pdf}
\bookmark[dest=sumOchsengartenObergut.1]{Ochsengarten-Obergut}
\includepdf[pages=-, link, linkname=sumGinzling]{Ginzling.pdf}
\bookmark[dest=sumGinzling.1]{Ginzling}
\includepdf[pages=-, link, linkname=sumRotholz]{Rotholz.pdf}
\bookmark[dest=sumRotholz.1]{Rotholz}
\includepdf[pages=-, link, linkname=sumWalchsee]{Walchsee.pdf}
\bookmark[dest=sumWalchsee.1]{Walchsee}
\includepdf[pages=-, link, linkname=sumKoessen]{Koessen.pdf}
\bookmark[dest=sumKoessen.1]{Koessen}
\includepdf[pages=-, link, linkname=sumInnervillgraten]{Innervillgraten.pdf}
\bookmark[dest=sumInnervillgraten.1]{Innervillgraten}
\includepdf[pages=-, link, linkname=sumMatreiinOsttirol]{MatreiinOsttirol.pdf}
\bookmark[dest=sumMatreiinOsttirol.1]{Matrei in Osttirol}
\includepdf[pages=-, link, linkname=sumStJohannimWalde]{StJohannimWalde.pdf}
\bookmark[dest=sumStJohannimWalde.1]{St.~Johann im Walde}
